Probabilities and propensities: conversion experiences in online grocery shopping

### 2017-07-21 log.rmd
got further into section E -- cutting down to model

>>>>

## Fri 21 Jul 2017 09:22:28 BST
- did so much work on this yesterday -- today I want it to go more easily; and I don't want to do so much work on it. 
- link everything in the presentation to the logistic regression model; that's the linchpin -- a massive conditional probability statement
        - TODO: add a bit about this model ... and think how Popper material helps here 
- grocery shopping in principle cannot be about personalization because shopping itself isn't. Thats a fundamental problem.
        - DONE - at least in part: update the theory of shopping to reflect this
        - TODO: exemplify it in a number of ways
- although I didn't do as much, paper felt better because I've managed to localize down on the logistic regression model; it helps me combat the personalization problem better. 
- reached the customer-product matrix;

### 2017-07-21 reconstruction.rmd
got further into section E -- cutting down to model

[TBA: THIS PARA could be moved down] The shelves of large contemporary supermarket are the endpoint of global logistic supply chains, in all their logistical, value-transforming and brand-mediated hypercomplexity (see [@Neilson_2012; @Tsing_2009]).  Groceries imply a planetary geography of agriculture, industry, transport, communication and financialisation animated by flows of labour and capital. Encounters between this hyper-complex commodity-geography and people, even in the familiar confines of a supermarket, are no simple matter, either for shoppers or for supermarket operators such as Tesco.

@@

The anthropologist  Daniel Miller has argued [@Miller_2012] that  shopping takes place as a negotiation of discrepancies between normative and actual social order (for instance, between ideals of health and commitments to thrift). What Miller suggests here is that ... [TBA]  It can also be seen  as a sacrificial ritual that re-instates social order in various ways.

@@

The shopping list, whether written on the back of an envelope, or saved as a list in an online grocery shopping system, lies at the intersection of logistical flows, infrastructural orderings, and  lively negotiations around actual and normative social orders.  Shopping lists, therefore  are intersectional ordering devices that encapsulates a universe of possible references, and a teeming multitude of propensities with an actual local order (in both senses of that term: an ordering and an imperative). As people shop, either by  trawling along aisles packed with thousands of products, or scrolling down screens or searching for particular brands amidst search results, lists inscribe some kind of order that filters or reduces the excessive propensities, claims, dazzle and distraction of commodities to the practicalities of domestic economy.

@@

Patel introduced the new 'personal relevance model' with a data graphic familiar to machine learners and statistical modellers. (A sketch of her graph appears in Figure \@ref(fig:models).)  The graph plots the _precision_ -- the proportion of the recommended products that customers actually purchase -- for several different statistical models.  The graph indexes the 'causal efficacy' of the recommender system, its capacity to include and transform propensities or 'real potentialities' into operational events or purchases.  The graph compares the previous rule-based recommender systems with some of the relevancy-model alternatives -- logistic regression, random forests, gradient boosting and a few others -- in terms of their predictions and how those predictions turned out. Patel dismissed most of the models quite quickly and focused only on one, the logistic regression model, which did as well or slight other than other alternatives.

The models Patel so briefly mentioned are all 'classifiers,' predictive models that 'classify' particular outcomes by calculating their probability.  Much of the core architecture of the machine learning classifier models received scant reference in Patel's presentation. It is most likely that in DunnHumby's work, the classifying model at the heart of the recommender system constructs a decision boundary in vector space between points that belong to different classes of things [@Mackenzie_2015e]. The classes of things are binary: `recommended` or `not recommended,` and for many of the models shown in the graph, membership of a thing such as grocery to the binary classes are expressed as a probability.  The logistic regression model generates probabilities for each product for each customer. The crucial point here is that the shopping list written by the customer is extended, quasi-invisibly, in the recommender system to potentially any of the 250,000 items in the Tesco inventory.

[HERE]

Patel did not articulate the implication of the contrast between the left and right hand side of the graphic. Although the statistical framing appears explicitly (e.g. in measuring the precision of  recommender systems), the underlying shift to a statistical model is barely mentioned.

@@

52 weeks of data
> in many kinds of events this is not the case, and the propensities cannot be measured because the relevant situation changes and cannot be repeated 17

>>>>

@@

The anthropologist  Daniel Miller has argued [@Miller_2012] that  shopping takes place as a negotiation of discrepancies between normative and actual social order (for instance, between ideals of health and commitments to thrift). What Miller suggests here is that shopping practices attempt to resolve difference between how people think they should live and how they actually live: 'we have to watch how shopping helps resolve these discrepancies between the normative and the actual, but we also need some ideas of where the normative comes from in the first place' [@Miller_2012, 72]. Whatever theory we have of how normative social order comes about, the key argument is that shopping inhabits a space between normative/idea and actual order. Importantly, and this will be a key consideration for recommender systems, shopping is not particularly personal or individual. Grocery shopping in particular intersects forms of social order concerning family and other forms of social grouping.

@@

The shopping list, whether written on the back of an envelope, or saved as a list in an online grocery shopping system, lies at the intersection of logistical flows, infrastructural orderings, and  lively negotiations around actual and normative social orders.[^30.1]  Shopping lists, therefore  are intersectional ordering devices that encapsulates a universe of possible references, and a teeming multitude of propensities with an actual local order (in both senses of that term: an ordering and an imperative). As people shop, either by  trawling along aisles packed with thousands of products, or scrolling down screens or searching for particular brands amidst search results, lists inscribe some kind of order that filters or reduces the excessive propensities, claims, dazzle and distraction of commodities to the practicalities of domestic economy.

[^30.1]: The shelves of large contemporary supermarket are the endpoint of global logistic supply chains, in all their logistical, value-transforming and brand-mediated hypercomplexity (see [@Neilson_2012; @Tsing_2009]).  Groceries imply a planetary geography of agriculture, industry, transport, communication and financialisation animated by flows of labour and capital. Encounters between this hyper-complex commodity-geography and people, even in the familiar confines of a supermarket, are no simple matter, either for shoppers or for supermarket operators such as Tesco.

@@

Patel introduced the new 'personal relevance model' with a data graphic familiar to machine learners and statistical modellers. (A sketch of her graph appears in Figure \@ref(fig:models).)  The graph plots the _precision_ -- the proportion of the recommended products that customers actually purchase -- for several different statistical models.  The graph indexes the 'causal efficacy' of the recommender system, its capacity to include and transform propensities or 'real potentialities' into operational events or purchases. Although she did not dwell on it, Patel used the graph to compare the previous rule-based recommender systems with some of the 'personal relevance model' alternatives -- logistic regression, random forests, gradient boosting and a few others -- in terms of their predictions and how those predictions turned out. Patel dismissed most of the models quite quickly and focused only on one, the logistic regression model, which did as well or slight other than other alternatives.

The models Patel so briefly mentioned are all 'classifiers,' predictive models that 'classify' particular events or outcomes by calculating their probability of membership of some class of event or outcome.  Much of the core architecture of the machine learning classifier models received scant reference in Patel's presentation. It is most likely that in DunnHumby's work, the classifier at the heart of the recommender system calculates for a given customer the most likely products to be purchased.  The classes of events are binary: `recommended` or `not recommended.` Products are allocated to one of these classes depending on their calculated probability. A probability greater than `0.5` typically would be `recommended.`  The logistic regression model generates probabilities of purchase for each product for each customer.

The crucial point here is that the shopping list written by the customer is extended, quasi-invisibly, in the recommender system to potentially any of the 250,000 items in the Tesco inventory. But the basis of this re-writing of the grocery list is an intricate conditional probability statement. Even if Patel assumed that the audience understood the working of the logistic regression model, the bulk of her presentation concerned the obstacles and problems that arise in trying to probabilise recommendations in ways that lead to the much-desired 'conversion events' or sales.

@@

I will not discuss in detail the different dimensions of probabilisation implicit in the architecture of the new model. It is worthwhile, however, highlighting some key elements that point directly to the troubled notion of 'personal relevance' and personalization more generally. These include the problem of repetition, attempts to create new views of the data, experiments on the platform, and the temporalities of calculation.

One fundamental difficulty here is that the propensity of a customer to buy a recommended product at a particular point in time is changing. Popper's notion of probabilities as propensities as tendencies to realization inherent in a situation directly implies this.  Measuring propensities and expressing these measurements as numbers between `0` and `1` becomes difficult because the situation is change and the propensities themselves change. As Popper observes, 'the propensities cannot be measured because the relevant situation changes and cannot be repeated' [@Popper_1990, 17]. No doubt the contingencies associated with grocery shopping are legion, perhaps more so than other kinds of products (books, music, films),  and any attempt to measure propensities associated with particular commodities  will encounter many dynamic changes.

One way in which recommender systems attempt to address the problem of the relational dynamics of propensities is by looking for repetitions. Patel mentioned that the new model uses  '52 weeks of data' for each customer. In including this data in the model, the assumption is that the probability that a specific customer will buy a specific product will be shaped by previous purchases. The history of previous purchases constitutes forms of repetition that imply a stabilisation of propensities (high or low; e.g. a vegan customer will never have purchased chicken products, so the measured propensities for any of the 1000 or so chicken products sold be Tesco will remain close to `0`).

When the logistic regression model includes the 52 weeks of previous purchases, the conditional probability calculation undertaken by the recommender system ramifies tremendously in several respects. Each of Tesco's 200,000 products becomes a variable in the classifier. Practically, most of the these variables will not influence the calculation of the probabilities for recommendations very much. As Patel observed,  again assuming that this would be obvious to the audience of data scientists,  'we have lots of zeros'.  -- points to the difficult terrain of  'good feature' thinking: the vast, nearly empty matrix of customer-product associations. Given that most things in Tesco remain relatively unknown to each other and to customers. A matrix that records associations between individual people and different products is bound to be mostly empty. Say Tesco has 1 million online customers. Each online  shopper has bought some selection of the 200,000 products. The customer-product data matrix will be `r format(1e6 * 2e5)` in size.  The product-customer matrix, the basic vector-space in which all recommender systems operate, remains very sparse and unpopulated. Given that any one customer is likely to only have bought 100 or so different products, the probability matrix will be close to `r format(100-(100/2e5)*100, 2)`% empty. The data is, as Patel put it, 'massively unbalanced' (any dataset where the items of interest are much rarer than some normal values is said to be unbalanced) and such imbalance would heavily bias the recommender system towards common and only impersonally relevant suggestions, suggestions that might  not produce the desired conversion experience.

HERE

```{r sparsity-ex, echo=FALSE, cache=TRUE, fig.margin=FALSE, fig.cap ='Sparsity of the customer-product association matrix' }
library(ggplot2)
library(reshape2)
p = as.logical(rbinom(n=2e5, size=20,0.001 ))
m = (matrix(p, nrow=1000, byrow=TRUE))
melted <- melt(m)
melted$value <- as.logical(melted$value)
ggplot(melted, aes(x = Var2, y = Var1, fill = value)) + geom_tile() +
    scale_fill_manual(values = c("grey", "red")) + xlab('Customer id') + ylab('Product id') + theme(legend.position='none')
```

Since so many people buy milk, the recommendation system might end up always recommending milk products. So the data needs to be 'corrected' by, as Patel reports, removing -- under-sampling -- some of the data for common purchases.   Secondly, even if a recommendation system finds uncommon items that are good recommendations, they should neither be too cheap ('low spend'), nor, as we will see too similar to what the customer has already ordered.
